Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-03-20 paper/tidy-missing-data-paper.Rmd
replace $r$ with $b$ - closes #12

The convention for representing missingness is a matrix, $R$, for data $y$ with $i$ rows and $j$ columns:

$$
r_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
$$

There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012. The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $r_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

The missingness matrix $R$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data. This is an idea upon which this new work builds.

@@

To explore missing data there needs to be a robust representation of missing
values that integrates into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The $R$ matrix, where 0 and 1
indicate not missing and missing provides a useful template, as it is very common in
missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The values 0 and 1 can
be confusing representations of missing values - does
0 indicate an absence of observation, or the presence of a missing value? 
The shadow matrix could be interactively linked to the data.

@@

The $R$ matrix is a very common representation of missing values. It can be created in the "R" language with \code{is.na(data)}. The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing. Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

@@

The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $R$. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

Common missing data operations can be considered verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

There are many common values for "missing". For example, "N/A", "MISSING", `-99`. Before replacing these values with missing values (`NA`), it is useful to _scan_ the data to understand the magnitude of the problem.  `miss_scan_count()` finds where specified values occur in data. For example, a scan of all occurrences of -99 is done with `miss_scan_count(data, search = -99)`, returning a table of occurrences of that value for each variable. `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters. They contain values like "n/a", "na", and -99, and are designed to be used with `miss_scan_count` to identify common missing values.

@@

dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                         1,   "A",   -100,
                         3,   "N/A", -99,
                         NA,  NA,    -98,
                         -99, "E",   -101,
                         -98, "F",   -1)

# knitr::kable(dat_ms, 
#              caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

@@

## replace: Replacing missing value label with another {#verbs-replace-with}

Once the magnitude of possible missing values has been assessed, these values need to be replaced with missing values. This can be done with `replace_with_na()`.  A dataset could have the values -99 in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))` For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

## add: Adding missingness summary variables {#verbs-add-cols}

@@

## shadow: Creating nabular data {#verbs-nabular}

@@

## flag: Describing different types of  missing values {#verbs-recode}

@@

## impute: Imputing values {#verbs-impute}

>>>>

The convention for representing missingness is a **b**inary matrix, $B$, for data $y$ with $i$ rows and $j$ columns:

$$
b_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
$$

There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

The missingness matrix $B$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data. This is an idea upon which this new work builds.

@@

To explore missing data there needs to be a robust representation of missing
values that integrates into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The $B$ matrix, where 0 and 1
indicate not missing and missing provides a useful template, as it is very common in
missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The values 0 and 1 can
be confusing representations of missing values - does
0 indicate an absence of observation, or the presence of a missing value? 
The shadow matrix could be interactively linked to the data.

@@

The $B$ matrix is a very common representation of missing values. It can be created in the "R" language with \code{is.na(data)}. The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing. Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

@@

The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $B$. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**. Data can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data matrix can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. 
The function is called  `miss_scan_count()`, and it returns a table of occurrences of that value for each variable. Two more data objects, `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters, to help users check for the typical representations of missing.

@@

dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                         1,   "A",   -100,
                         3,   "N/A", -99,
                         NA,  NA,    -98,
                         -99, "E",   -101,
                         -98, "F",   -1)
```

@@

### replace: Replacing missing value label with another {#verbs-replace-with}

Once the different representations of possible missing values have been identified, these values can be replaced using the function, `replace_with_na()`.  A dataset could have the values `-99` in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))`. For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

### add: Adding missingness summary variables {#verbs-add-cols}

@@

### shadow: Creating nabular data {#verbs-nabular}

@@

### flag: Describing different types of  missing values {#verbs-recode}

@@

### impute: Imputing values {#verbs-impute}
